---
title: "Data Analysis"
author: "Ina Liao"
date: "2024-02-05"
output: pdf_document
---


```{r Setup, include=FALSE}
knitr:::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=FALSE,fig.align = "center", dev = "cairo_pdf", fig.pos = "H")
```

```{r Install packages, message=FALSE}
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("forecast")
#install.packages("here")
#install.packages("knitr")
#install.packages("kableExtra")
#install.packages("dplyr")
#install.packages("openxlsx")
#install.packages("ggthemes")
#install.packages("tidyr")
#install.packages("GGally")
#install.packages("tseries")
#install.packages("blorr")
library(lubridate)
library(ggplot2)
library(forecast) 
library(here)
library(knitr)
library(kableExtra)
library(dplyr)
library(openxlsx)
library(ggthemes)
library(tidyr)
library(GGally)
library(trend)
library(tseries)
library(blorr)   
library(lmtest) 
```

```{r Import Data, echo=TRUE, results='hide'}
raw_DUK<-read.csv(here('Data/Raw/load_DUK.csv'),stringsAsFactors = TRUE,skip=0,header=TRUE)
head(raw_DUK)
#column 45,46,47,48,49,50,51,52,53: electricity transmitted from other ISO/RTO

raw_weather<-read.csv(here("Data/Processed Data/Processed_weather_2012-2022.xlsx"),stringsAsFactors = TRUE,skip=0,header=TRUE)
head(raw_weather)
```

```{r Selected the needed data - Load}
df_DUK<-raw_DUK[,c(3,5,8,9,10,11)]
head(df_DUK)
colnames(df_DUK)

#change the date format as ymd
df_DUK$Local.date<-dmy(df_DUK$Local.date)

#combine date and time
df_DUK$Local.time<-as.POSIXct(df_DUK$Local.time,format = "%d%b%Y %H:%M")
df_DUK$Local.time<-format(df_DUK$Local.time,"%Y-%m-%d %H:%M")

#rename column names 
colnames(df_DUK)[colnames(df_DUK)=="DF"]<-"Demand Forecast"
colnames(df_DUK)[colnames(df_DUK)=="D"]<-"Demand"
colnames(df_DUK)[colnames(df_DUK)=="NG"]<-"Net Generation"
colnames(df_DUK)[colnames(df_DUK)=="TI"]<-"Total Interchange"

#set as numaric
for (i in 3:6){
  df_DUK[,i]<-as.numeric(df_DUK[,i])
}

#mutate forecast error 
df_DUK<-df_DUK %>%
  mutate(Forecast.Error=`Demand Forecast`- Demand)

#mutate calendar event
df_DUK$day_of_week <- wday(df_DUK$Local.date, label = FALSE)
df_DUK<-df_DUK[,c(1:2,8,3:7)]

head(df_DUK)
```

```{r Select the needed data - Weather}
#change the date format as ymd
raw_weather$Date<-paste(raw_weather$Year,raw_weather$Month,raw_weather$Day,sep="-")
raw_weather$Date<-ymd(raw_weather$Date)

#set the time format as hour:minute
raw_weather$Time<-paste(raw_weather$Hour,raw_weather$Minute,sep=":")
raw_weather$Time<-as.POSIXct(raw_weather$Time, format = "%H:%M")
raw_weather$Time<-format(raw_weather$Time,"%H:%M")

#combine date and time
raw_weather$Time2<-paste(raw_weather$Date,raw_weather$Time, sep=" ")
df_weather<-raw_weather[,c(9,11,7,8)]
head(df_weather)

#mutate calendar event
df_weather$day_of_week<-wday(df_weather$Date, label = FALSE)

#reorder dataframe
df_weather<-df_weather[,c(1:2,5,3:4)]
```


```{r Missing Data - Load}
#create time dataframe
start_time_DUK<-as.POSIXct("2015-07-01 01:00:00")
end_time_DUK<-as.POSIXct("2023-10-05 00:00:00")
df_time_DUK<-data.frame(Local.time=seq(from=start_time_DUK, to=end_time_DUK, by="1 hour"))
df_time_DUK$Local.time<-format(df_time_DUK$Local.time,"%Y-%m-%d %H:%M")

#merge dataframes
df_DUK_merge<-left_join(df_time_DUK,df_DUK,by="Local.time")
head(df_DUK_merge)

#check if there is missing data
df_DUK_merge %>%
  filter(is.na(Local.time)) 
#no missing data
```

```{r Missing Data - Weather}
#create time dataframe
start_time_weather<-as.POSIXct("2012-01-01 0:30")
end_time_weather<-as.POSIXct("2022-12-31 23:30")
df_time_weather<-data.frame(Time2=seq(from=start_time_weather, to=end_time_weather, by="1 hour"))
df_time_weather$Time2<-format(df_time_weather$Time2,"%Y-%m-%d %H:%M")

#merge dataframes
df_weather_merge<-left_join(df_time_weather,df_weather,by="Time2")
head(df_weather_merge)

#check if there is missing data
df_weather_merge %>%
  filter(is.na(Date)) 
#missing the data on 2012-02-29, 2016-02-29, 2020-02-29
```

```{r Set Plot Theme}
my_plot_theme<- theme_calc()+
  theme(
    #plot background
    plot.background = element_rect(color="gray"),
    
    #plot title
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    
    #gridlines 
    panel.grid.major=element_line("white"),
    panel.grid.minor = element_line("white"),
    axis.ticks=element_line(color="gray"),
    
    #legend 
    legend.key=element_rect(color="white"),
    legend.background = element_rect(color="white"),
    legend.text = element_text(size = 8),
    legend.position="right"
      )
theme_set(my_plot_theme)
```

```{r Daily Average - Load}
#can't merge two datasets by time since the recorded hours are different (i.e., 01:00 and 01:30)
#calculate the daily average then merge the dataframes

#the outliers haven't been removed
#the missing value will be removed when computing the mean

df_daily_DUK<-df_DUK_merge %>%
  
  #calculate daily average
  group_by(Local.date,day_of_week) %>%
  summarize(
    demand_forecast=mean(`Demand Forecast`, na.rm=TRUE),
    demand=mean(Demand, na.rm=TRUE),
    net_generation=mean(`Net Generation`, na.rm=TRUE),
    total_interchange=mean(`Total Interchange`, na.rm=TRUE),
  ) %>%
  
  #mutate forecast error 
  mutate(forecast_error=demand_forecast- demand)
head(df_daily_DUK)
```

```{r Daily Average - Weather}
df_daily_weather<-df_weather_merge %>%
  group_by(Date,day_of_week) %>%
  summarize(
    dew_point=mean(Dew.Point, na.rm=TRUE),
    wind_speed=mean(Wind.Speed, na.rm=TRUE)
  ) 
head(df_daily_weather)
```

```{r Merge Two Dataframe}
colnames(df_daily_DUK)[colnames(df_daily_DUK) == "Local.date"] <- "Date"
df_all<-left_join(df_daily_weather,df_daily_DUK,by="Date")

#weather and load data have different starting and ending date 
#weather data: 2012-01-01 - 2022-12-31
#load data: 2015-07-01 - 2023-10-04

#trim the dataframe
start_date<-ymd(first(df_daily_DUK$Date))
#end_date<-ymd(last(df_daily_weather$Date))  #return NA

#create another variable to identify weekends and weekdays
df<-df_all %>%
  filter(Date>=start_date) %>%
  mutate(weekday_weekend=ifelse(day_of_week.x %in% 1:5,0,1)
    )
df<-df[,c(1:2,11,3:4,6:10)]
head(df)
tail(df) #last day is 2022-12-31	
```

```{r Time Series Object - TS}
#ts
ts_all<-ts(df[,4:10], start=c(2015,07),end=c(2022,12),frequency=365) #correct frequency?
autoplot(ts_all)

#electricity 
plot_ts_electricity<-autoplot(ts_all[,3:7])+
  labs(y="electicity (MWh)")
plot_ts_electricity

#weather
plot_ts_weather<-autoplot(ts_all[,1:2])
plot_ts_weather

plot_ts_electricity_weather<-autoplot(ts_all[,4],series="electricity demand")+
  autolayer(ts_all[,7],series="forecast error")+
  autolayer(ts_all[,1],series="dew point")+
  autolayer(ts_all[,2],series="wind speed")
plot_ts_electricity_weather

#remove outliers?
#demand froecast is around 1-2 MWh: 2018-03-11, 2019-03-10, 2018-04-05
#demand, net generation, and total interchange are around the same level in 2018-03-11, 2019-03-10

```

```{r Time Series Object - MSTS}
#msts
msts_all<-msts(df[,4:10], seasonal.periods=12, start=c(2015,07),end=c(2022,12))

#electricity 
plot_msts_electricity<-autoplot(msts_all[,3:7])+
  labs(y="electicity (MWh)")
plot_msts_electricity

#weather
plot_msts_weather<-autoplot(msts_all[,1:2])
plot_msts_weather

plot_msts_electricity_weather<-autoplot(msts_all[,4],series="electricity demand")+
  autolayer(msts_all[,7],series="forecast error")+
  autolayer(msts_all[,1],series="dew point")+
  autolayer(msts_all[,2],series="wind speed")
plot_msts_electricity_weather

```

```{r ACF and PACF}
#place plot side by side
#par(mfrow=c(7,2)) 
#error: Error in plot.new() : figure margins too large

for(i in 1:7){ 
  #plot(ts_all$Date,ts_all[,i],type="l",col="black",
       #ylab=NA,xlab=NA,main=paste(colnames(ts_all[i]),sep=" ")) #error: the length of df and ts are different 
  Acf(ts_all[,i],lag.max=40,main=paste("ACF for",colnames(df[,i+3]),sep=" "),ylim=c(-0.5,1))
  Pacf(ts_all[,i],lag.max=40,main=paste("PACF for",colnames(df[,i+3]),sep=" "),ylim=c(-0.5,1))
}
```
1. Dew point: does not have significant trend but have a drop in 2022; has a strong time-dependent (correlation still high in the higher lag); might not have seasonality 
2. Wind speed: does not have significant trend; might have seasonal components
3. Demand, demand forecast, and net generation might have an increased trend and have seasonal components
4. Demand, forecast demand, forecast error, and net generation have a peak in late 2022
 


```{r Stationary test}
#seasonal Mann-Kendall
#season 1-10?
smk_dew<-summary(trend::smk.test(ts_all[,1])) 
smk_wind<-summary(trend::smk.test(ts_all[,2])) 
smk_demandforecast<-summary(trend::smk.test(ts_all[,3]))
smk_demand<-summary(trend::smk.test(ts_all[,4]))
smk_generation<-summary(trend::smk.test(ts_all[,5]))
smk_interchange<-summary(trend::smk.test(ts_all[,6]))
smk_forecast_error<-summary(trend::smk.test(ts_all[,7]))

#ADF test
ADF_dew<-print(adf.test(ts_all[,1],alternative = "stationary"))
ADF_wind<-print(adf.test(ts_all[,2],alternative = "stationary"))
ADF_demandforecast<-print(adf.test(ts_all[,3],alternative = "stationary"))
ADF_demand<-print(adf.test(ts_all[,4],alternative = "stationary"))
ADF_generation<-print(adf.test(ts_all[,5],alternative = "stationary"))
ADF_interchange<-print(adf.test(ts_all[,6],alternative = "stationary"))
ADF_forecast_error<-print(adf.test(ts_all[,7],alternative = "stationary"))
```

From Seasonal Mann-Kendall test, all p-values are greater than 0.05, and thus we do not have enough evidence to reject the null hypothesis that the series is stationary. From the Augmented Dickey-Fuller (ADF) test, all p-values are smaller than 0.05, meaning that we have sufficient evident to support that the series do not have stochastic trend. (the result from two tests all refer that the series are stationary)

```{r Mitigate Autocorrelation - Linear}
decompose_dew<-stl(ts_all[,1],s.window="periodic") #is s.window argument correct?
decompose_wind<-stl(ts_all[,2],s.window="periodic")
decompose_demandforecast<-stl(ts_all[,3],s.window="periodic")
decompose_demand<-stl(ts_all[,4],s.window="periodic")
decompose_generation<-stl(ts_all[,5],s.window="periodic")
decompose_interchange<-stl(ts_all[,6],s.window="periodic")

plot(decompose_dew,main=paste(colnames(df[,1+3])))
plot(decompose_wind,main=paste(colnames(df[,2+3])))
plot(decompose_demandforecast,main=paste(colnames(df[,3+3])))
plot(decompose_demand,main=paste(colnames(df[,4+3])))
plot(decompose_generation,main=paste(colnames(df[,5+3])))
plot(decompose_interchange,main=paste(colnames(df[,6+3])))
```

```{r Mitigate Autocorrelation - Differencing}
#given that all the series are not stochastic trend, is it necessary to use differencing method to detrend?
#using first order approach

ts_diff<-diff(ts_all[,1:7],lag = 1,differences = 1)

#plot original trend and differenced trend
#autoplot does not work in for loop 
plot_dew_diff<-autoplot(ts_diff[,1],series="Differenced Series")+
    autolayer(ts_all[,1],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,1+3]),sep="_"))
plot_dew_diff

plot_wind_diff<-autoplot(ts_diff[,1],series="Differenced Series")+
    autolayer(ts_all[,2],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,2+3]),sep="_"))
plot_wind_diff

plot_deamndforecast_diff<-autoplot(ts_diff[,3],series="Differenced Series")+
    autolayer(ts_all[,3],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,3+3]),sep="_"))
plot_deamndforecast_diff

plot_deamnd_diff<-autoplot(ts_diff[,4],series="Differenced Series")+
    autolayer(ts_all[,4],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,4+3]),sep="_"))
plot_deamnd_diff

plot_generation_diff<-autoplot(ts_diff[,5],series="Differenced Series")+
    autolayer(ts_all[,5],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,5+3]),sep="_"))
plot_deamnd_diff

plot_interchange_diff<-autoplot(ts_diff[,6],series="Differenced Series")+
    autolayer(ts_all[,6],series="Time Series")+
    labs(x=NA,color="",title=paste(colnames(df[,6+3]),sep="_"))
plot_interchange_diff

#for(i in 1:4){
  #plot(ts_diff[,i],type="l",col="black",
       #ylab=NA,xlab=NA,main=paste(colnames(df[,i+3]),sep="_"))+
     #plot(ts_all[,i],type="l",col="pink",
       #ylab=NA,xlab=NA,main=paste(colnames(df[,i+3]),sep="_"))
```
After differencing the series, the trend level of dew points has been decreased. 

```{r Multilinear Regression}
head(df)
model_all<-lm(demand~wind_speed+dew_point+weekday_weekend+demand_forecast+net_generation+total_interchange,data=df)
summary(model_all)

model_factors<-lm(demand~wind_speed+dew_point+weekday_weekend,data=df)
summary(model_factors)

#linear regression: can we assume that independent variables are normally distributed
```
1. On average, weekend that falls above the data set has an 292.172 greater electricity demand than weekday while holding other explanatory variables constant. The slope parameter is significantly different than zero at a significance level of 0.05 (p-value<0.05).
2. 7.3% of the variation in the electricity demand is explained by the model, indicating that the model may not be a good fit for predicting electricity demand.

```{r Residuals vs fitted}
vif(model_all)
vif(model_factors)
```

There might have a multicollinearity occur between variables since all variables have relatively low variation inflation factor.